---
title: "DA-CV"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{DA-CV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
Dissimilarity-adaptive cross-validation (DA-CV) was developed by @Wang2025. 
It uses adversial validation (AV) to predict the probability that a prediction location is similar to the training samples.
Then it calculates a RMSE based on random CV, as well as spatial+ CV @Wang2023 and weights both of them using the relative area 
of similar cells (random CV) and dissimilar cells (spatial+ CV). 

The more detailed workflow is:

1. Generate an AV dataset consisting of the training samples and an equal number of randomly sampled prediction locations.
2. Train the AV classifier based on the available predictors to predict the probability of a data point to belong to the training samples using 50% of the AV dataset.
3. Evaluate the AV classifier on the other 50% and calcuate the AUC.
4. Normalize the AUC to [0,1] by starting with 0 at $AUC = 0.5$. This normalized AUC is then interpreted as the overall dissimilarity $D$.
5. Derive the threshold to distinguish similar from dissimilar areas from the overall dissimilarity as $T(D) = 0.5*D$.
6. Predict the similarity of all prediction locations using the AV classifier.
7. Binarize the similarity map using $T(D)$.
8. Calculate the weighted RMSE as $RMSE_{DA} = \sqrt{W_{RDM} * RMSE^2_{RDM} + W_{SP} * RMSE^2_{SP}}$ 

Spatial+ CV works as follows:

1. Divide samples into blocks using agglomerative hierarchical clustering, where the maximum linkage (i.e., the maximum distance between samples in one cluster / block size) is derived from the semivariogram of the response values measured at the sampling locations (i.e., spatial autocorrelation range)
2. Average the predictor values, response values and coordinates of all samples belonging to one block.
3. Cluster the blocks in three ways:

	3.1. K-Means clustering of the coordinates

	3.2. K-Prototypes clustering of the predictor values
	
	3.3. K-Means clustering of the response

4. Combine the three clustering results into k folds using the cluster ensemble function "Hybrid Bipartite Graph Formulation" that finds consistency between the three clustering results.


## Setup
```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
```

```{r setup}
library(PDAV)
library(terra)
library(sf)
library(simsam)
library(ggplot2)
library(tidyterra)
library(ggnewscale)
library(dplyr)
```

## Example data
### Simulate predictors and response
```{r}
set.seed(100)
rast_grid = terra::rast(xmin = 0, xmax = 200, ymin = 0, ymax = 200, ncols = 200, nrows = 200)
r <- sim_covariates(rast_grid, vgm = gstat::vgm(model = "Exp", psill = 1, range = 20), n = 4)
outcome <- simsam::blend_rasters(r, ~ cov1 + cov2 + cov3 + cov4)
r <- c(outcome, r)
terra::crs(r) <- "EPSG:3857"

response_name <- "outcome"
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| echo: false

ggplot() +
	geom_spatraster(data = r) +
	coord_sf(expand = FALSE, datum = st_crs(r)) +
	scale_fill_viridis_c() +
	facet_wrap(~lyr) +
	theme_bw() +
	theme(panel.grid = element_blank())
```

### Simulate samples
```{r}
library(gstat)
library(sp)
sample_random <- sam_field(
	x = r,
	size = 100,
	type = "random",
	na.rm = TRUE
) |>
	terra::extract(x = r, bind = TRUE) |>
	st_as_sf()


# Clustered sampling
set.seed(456)
sample_clustered <- sam_field(
	x = r,
	size = 100,
	type = "clustered",
	type_opts = list(nclusters = 5, radius = 15),
	na.rm = TRUE
) |>
	terra::extract(x = r, bind = TRUE) |>
	st_as_sf()
```

```{r semivariogram}
#| eval: false
#| echo: false

sample_random_sp <- as(sample_random, "Spatial")
sample_clustered_sp <- as(sample_clustered, "Spatial")
vg_random <- variogram(as.formula(paste(response_name, "~ 1")), data = sample_random_sp)
vg_clustered <- variogram(as.formula(paste(response_name, "~ 1")), data = sample_clustered_sp)
plot(vg_random, main = paste("Semivariance of", response_name, "(Random samples)"))
plot(vg_clustered, main = paste("Semivariance of", response_name, "(Clustered samples)"))
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| echo: false

samples <- rbind(sample_random |> mutate(sampling = "random"), sample_clustered |> mutate(sampling = "clustered")) |>
	mutate(sampling = factor(sampling, levels = c("random", "clustered")))

ggplot() +
	geom_spatraster(data = r$outcome) +
	scale_fill_viridis_c("outcome") +
	new_scale_fill() +
	geom_sf(data = samples, shape = 21, aes(fill = "training samples"), size = 1) +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	scale_fill_manual("", values = "white") +
	facet_wrap(~sampling) +
	theme_bw() +
	theme(panel.grid = element_blank())
```

## Run DA-CV

Now, we use DA-CV to obtain area estimates of interpolation vs extrapolation areas, that can then be used to derive a weighted RMSE. We leave the autocorrelation threshold at 1, despite the sample variogram showing indicating a range of 5-10, because higher autocorrelation thresholds would result in spatial+ CV-splits that has a smaller size than the sample size. Maybe due to empty clusters?

```{r}
result_random <- da_cv(
	samples = sample_random,
	predictors = r,
	response = response_name,
	folds_k = 5,
	autoc_threshold = 1,
	cate_num = 0
)

# Using clustered samples
result_clustered <- da_cv(
	samples = sample_clustered,
	predictors = r,
	response = response_name,
	folds_k = 5,
	autoc_threshold = 1,
	cate_num = 0
)

```

For randomly distributed samples, the AV classifier has a performance of $AUC = `r round(result_random$auc, 1)`$. This is then normalized to $D = 0$, because the classifier is not better than randomly guessing if a prediction location is similiar or dissimilar (see @Wang2025b, section 2.1 for more information). The threshold is then calculated as $T(D) = 0.5 * 0$, which is 0. This means that all prediction locations are similar to the training 1, and no extrapolation is required, as can be seen in the following map. All weights will be given to random CV. 

For clustered samples, the AV classifier achieves a performance of $AUC = `r round(result_clustered$auc, 2)`$. This leads to $D = \frac{`r round(result_clustered$auc, 2)` - 0.5}{1 - 0.5} = `r (round(result_clustered$auc, 2) - 0.5) / 0.5`$. The threshold is then $T(D) = 0.5* `r (round(result_clustered$auc, 2) - 0.5) / 0.5` = `r 0.5*((round(result_clustered$auc, 1) - 0.5) / 0.5)`$. Hence, all prediction cells with a similarity score lower than `r 0.5*((round(result_clustered$auc, 1) - 0.5) / 0.5)` are classified as dissimilar. The relative fraction of prediction locations being similar from the sampling locations is `r round(result_clustered$weights[["similar"]], 2)`, while the fraction being dissimilar is `r round(result_clustered$weights[["different"]], 2)`.

``` {r}
#| fig-width: 6
#| fig-height: 4

plot(result_random) +
	new_scale_fill() +
	geom_sf(data = sample_random, shape = 21, aes(fill = "sampling locations")) +
	coord_sf(expand = FALSE, datum = st_crs(sample_random)) +
	scale_fill_manual("", values = "white") +
	theme(legend.position = "right")

plot(result_clustered) +
	new_scale_fill() +
	geom_sf(data = sample_clustered, shape = 21, aes(fill = "sampling locations")) +
	coord_sf(expand = FALSE, datum = st_crs(sample_clustered)) +
	scale_fill_manual("", values = "white") +
	theme(legend.position = "right")

```

Lastly, the resulting cross-validation fold assignments could be used to calculate the weighted RMSE of a spatial predictive model.

```{r}
#| fig-width: 8
#| fig-height: 6

sample_random$randomCV <- result_random$folds_RDM
sample_random$spatialCV <- result_random$folds_SP
sample_clustered$randomCV <- result_clustered$folds_RDM
sample_clustered$spatialCV <- result_clustered$folds_SP

samples <- rbind(sample_random |> mutate(sampling = "random"), sample_clustered |> mutate(sampling = "clustered")) |>
	select(randomCV, spatialCV, sampling)
samples$ID <- 1:nrow(samples)
samples_geom <- select(samples, ID)
samples_long <- st_drop_geometry(samples) |>
	pivot_longer(-c(ID, sampling)) |>
	left_join(samples_geom, by = "ID") |>
	st_as_sf() |>
	mutate(
		weight = case_when(
			sampling == "random" & name == "randomCV" ~ round(result_random$weights[["similar"]], 1),
			sampling == "random" & name == "spatialCV" ~ round(result_clustered$weights[["similar"]], 1),
			sampling == "clustered" & name == "randomCV" ~ round(result_random$weights[["different"]], 1),
			sampling == "clustered" & name == "spatialCV" ~ round(result_clustered$weights[["different"]], 1)
		),
		sampling = factor(sampling, levels = c("random", "clustered"))
	)

ggplot() +
	geom_sf(data = samples_long, aes(colour = as.factor(value)), size = 1) +
	geom_text(
		data = samples_long,
		aes(x = 25, y = 180, label = paste0("Weight: ", weight)),
		size = 3,
		check_overlap = TRUE
	) +
	coord_sf(expand = FALSE, datum = st_crs(sample_clustered)) +
	facet_grid(name ~ sampling) +
	scale_color_brewer("Fold", palette = "Set1") +
	theme_bw() +
	theme(panel.grid = element_blank(), legend.position = "right") +
	labs(x = "", y = "")

```