---
title: "k-NNDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k-NNDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Introduction
k-fold nearest-neighbour distance matching (kNNDM) cross-validation was developed by @Linnenbrink2024.
It is the k-fold version of NNDM (@Mila2022), and as such much faster and better suited towards medium to large sized datasets.
As NNDM, kNNDM also matches the distribution of nearest-neighbour distances (NND) between prediction locations and training samples during CV, but opposed to NNDM uses a clustering algorithm instead of excluding training samples.

The more detailed workflow is:

1. Conmpute the distribution of NNDs between training samples ($G_j$), as well as between prediction locations and training samples ($G_{ij}$).
2. Use the KS-test to evaluate wether $G_j >= G_{ij}$, i.e., wether the samples are randomly distributed in the prediction area.
3. A) If $G_j >= G_{ij}$; return a random CV split and stop the algorithm here.
3. B) Otherwise, continue by clustering the training points into $q \in Q$, where $Q = [k, ..., N]$. This yields a gradient from $k$ clusters (i.e., one cluster per fold) to $N$ clusters (i.e., one cluster per training sample).
4. The $q$ clusters are then merged along the first principal component of the training points coordinates' until $k$ is reached. This means that for $q = N$, nothing happens and a strongly clustered fold configuration is returned. For $q = N$, a random fold assignment is returned. For everything in-between ($q>k<N$), a gradient from strongly to weak clustered folds is returned.
5.  For each of the $q$ fold configurations, the NND distribution between the unique folds is calculated ($G_{j}^*$), and the distance between $G_j^*$ and $G_{ij}$ is calculated using the Wasserstein statistic (W).
6. The fold configuration yielding the lowest W is returned.


## Setup
```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
```

```{r setup}
library(PDAV)
library(terra)
library(sf)
library(simsam)
library(ggplot2)
library(tidyterra)
library(ggnewscale)
library(dplyr)
```

## Example data
### Simulate predictors and response
```{r}
set.seed(100)
rast_grid = terra::rast(xmin = 0, xmax = 200, ymin = 0, ymax = 200, ncols = 200, nrows = 200)
r <- sim_covariates(rast_grid, vgm = gstat::vgm(model = "Exp", psill = 1, range = 20), n = 4)
outcome <- simsam::blend_rasters(r, ~ cov1 + cov2 + cov3 + cov4)
r <- c(outcome, r)
terra::crs(r) <- "EPSG:3857"

response_name <- "outcome"
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| echo: false

ggplot() +
	geom_spatraster(data = r) +
	coord_sf(expand = FALSE, datum = st_crs(r)) +
	scale_fill_viridis_c() +
	facet_wrap(~lyr) +
	theme_bw() +
	theme(panel.grid = element_blank())
```

### Simulate samples
```{r}
sample_random <- sam_field(
	x = r,
	size = 100,
	type = "random",
	na.rm = TRUE
) |>
	terra::extract(x = r, bind = TRUE) |>
	st_as_sf()


# Clustered sampling
set.seed(456)
sample_clustered <- sam_field(
	x = r,
	size = 100,
	type = "clustered",
	type_opts = list(nclusters = 5, radius = 15),
	na.rm = TRUE
) |>
	terra::extract(x = r, bind = TRUE) |>
	st_as_sf()
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| echo: false

samples <- rbind(sample_random |> mutate(sampling = "random"), sample_clustered |> mutate(sampling = "clustered")) |>
	mutate(sampling = factor(sampling, levels = c("random", "clustered")))

ggplot() +
	geom_spatraster(data = r$outcome) +
	scale_fill_viridis_c("outcome") +
	new_scale_fill() +
	geom_sf(data = samples, shape = 21, aes(fill = "training samples"), size = 1) +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	scale_fill_manual("", values = "white") +
	facet_wrap(~sampling) +
	theme_bw() +
	theme(panel.grid = element_blank())
```

## kNNDM CV

```{r}

```