---
title: "k-NNDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k-NNDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Introduction
k-fold nearest-neighbour distance matching (kNNDM) cross-validation was developed by @Linnenbrink2024.
It is the k-fold version of NNDM (@Mila2022), and as such much faster and better suited towards medium to large sized datasets.
As NNDM, kNNDM also matches the distribution of nearest-neighbour distances (NND) between prediction locations and training samples during CV, but opposed to NNDM uses a clustering algorithm instead of excluding training samples.

The more detailed workflow is:

1. Conmpute the distribution of NNDs between training samples ($G_j$), as well as between prediction locations and training samples ($G_{ij}$).
2. Use the KS-test to evaluate wether $G_j >= G_{ij}$, i.e., wether the samples are randomly distributed in the prediction area.
3. A) If $G_j >= G_{ij}$; return a random CV split and stop the algorithm here.
3. B) Otherwise, continue by clustering the training points into $q \in Q$, where $Q = [k, ..., N]$. This yields a gradient from $k$ clusters (i.e., one cluster per fold) to $N$ clusters (i.e., one cluster per training sample).
4. The $q$ clusters are then merged along the first principal component of the training points coordinates' until $k$ is reached. This means that for $q = N$, nothing happens and a strongly clustered fold configuration is returned. For $q = N$, a random fold assignment is returned. For everything in-between ($q>k<N$), a gradient from strongly to weak clustered folds is returned.
5.  For each of the $q$ fold configurations, the NND distribution between the unique folds is calculated ($G_{j}^*$), and the distance between $G_j^*$ and $G_{ij}$ is calculated using the Wasserstein statistic (W).
6. The fold configuration yielding the lowest W is returned.


## Setup
```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
```

```{r setup}
library(PDAV)
library(terra)
library(sf)
library(simsam)
library(ggplot2)
library(cowplot)
library(tidyterra)
library(ggnewscale)
library(dplyr)

set.seed(100)

k = 2
```

## Example data
### Simulate predictors and response
```{r}
r <- PDAV:::generate_rast()
samples <- PDAV:::generate_samples(r, 100) |>
	filter(sampling == "biased")
samples$point_id <- 1:nrow(samples)
samples <- select(samples, point_id)
```

```{r}
#| fig-width: 8
#| fig-height: 4
#| echo: false

ggplot() +
	geom_spatraster(data = r) +
	coord_sf(expand = FALSE, datum = st_crs(r)) +
	scale_fill_viridis_c() +
	facet_wrap(~lyr) +
	theme_bw() +
	theme(panel.grid = element_blank())

ggplot() +
	geom_spatraster(data = r$outcome) +
	scale_fill_viridis_c("outcome") +
	new_scale_fill() +
	geom_sf(data = samples, shape = 21, aes(fill = "training samples"), size = 2) +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	scale_fill_manual("", values = "white") +
	theme_bw() +
	theme(panel.grid = element_blank())
```


## kNNDM in geographical space


### 1. Compute $G_j$ and $G_{ij}$

Firstly, the kNNDM algorithm calcuates the distribution of NND between samples ($G_j$), as well as the distribution of NNDs between prediction points and samples ($G_{ij}$). The calculation of NNDs is shown in the following figure: the nearest neighbour distance between samples is shown in the left panel, while the right panel shows nearest neighbour distances between prediction points and samples.

```{r}
#| fig-width: 8
#| fig-height: 4
#| echo: false

prediction_points <- spatSample(r, size = 1000, method = "regular", as.points = TRUE) |>
	st_as_sf()

pred_coords <- st_coordinates(prediction_points)
sample_coords <- st_coordinates(samples)

# Gj
idx_nn_s <- st_nearest_feature(samples)
nearest_samples_s <- samples[idx_nn_s, ]
samp_coords_s <- st_coordinates(nearest_samples_s)
lines_s <- mapply(
	function(i) {
		st_linestring(rbind(sample_coords[i, ], samp_coords_s[i, ]))
	},
	seq_len(nrow(sample_coords)),
	SIMPLIFY = FALSE
)
lines_sf_s <- st_sf(
	geometry = st_sfc(lines_s, crs = st_crs(prediction_points))
)

# Gij
idx_nn <- st_nearest_feature(prediction_points, samples)
nearest_samples <- samples[idx_nn, ]
samp_coords <- st_coordinates(nearest_samples)
lines <- mapply(
	function(i) {
		st_linestring(rbind(pred_coords[i, ], samp_coords[i, ]))
	},
	seq_len(nrow(pred_coords)),
	SIMPLIFY = FALSE
)
lines_sf <- st_sf(
	geometry = st_sfc(lines, crs = st_crs(prediction_points))
)


NND_samples <- ggplot() +
	geom_sf(data = prediction_points, alpha = 0) +
	geom_sf(data = samples, shape = 4) +
	geom_sf(data = lines_sf_s, linewidth = 0.2, color = "black") +
	scale_shape_manual("", values = c(1, 4)) +
	coord_sf(expand = FALSE, datum = st_crs(prediction_points)) +
	ggtitle("NND between samples") +
	theme_bw() +
	theme(legend.position = "None", panel.grid = element_blank(), plot.title = element_text(size = 12))

NND_pred_samples <- ggplot() +
	geom_sf(data = prediction_points, aes(shape = "prediction points"), color = "lightblue") +
	geom_sf(data = samples, aes(shape = "samples")) +
	geom_sf(data = lines_sf, linewidth = 0.2, color = "lightblue") +
	scale_shape_manual("", values = c(1, 4)) +
	coord_sf(expand = FALSE, datum = st_crs(prediction_points)) +
	ggtitle("NND between prediction points and samples") +
	theme_bw() +
	theme(legend.position = "None", panel.grid = element_blank(), plot.title = element_text(size = 12))

plot_grid(NND_samples, NND_pred_samples, nrow = 1, align = "vh")
```

To characterize the distribution of these NNDs, their empirical cumulative density functions (ECDFs) are calculated:

```{r}
Gj <- c(FNN::knn.dist(sample_coords, k = 1))
Gij <- c(FNN::knnx.dist(
	query = pred_coords,
	data = sample_coords,
	k = 1
))
```

```{r}
#| fig-width: 4
#| fig-height: 4
#| echo: false
#| warning: false

myColors <- RColorBrewer::brewer.pal(3, "Dark2")

Gij_df <- data.frame(r = Gij[order(Gij)])
Gij_df$Function <- "1_Gij(r)"
Gj_df <- data.frame(r = Gj[order(Gj)])
Gj_df$Function <- "3_Gj(r)"

# Merge data for plotting
Gplot <- rbind(Gij_df, Gj_df)

ggplot(
	data = Gplot,
	aes_string(x = "r", group = "Function", col = "Function")
) +
	geom_vline(xintercept = 0, lwd = 0.1) +
	geom_hline(yintercept = 0, lwd = 0.1) +
	geom_hline(yintercept = 1, lwd = 0.1) +
	stat_ecdf(geom = "step", lwd = 0.8) +
	scale_colour_manual(
		values = c(myColors[2], myColors[1]),
		labels = c("prediction-to-sample", "sample-to-sample")
	) +
	guides(colour = guide_legend(override.aes = list("linewidth" = 1.6))) +
	theme_bw(base_size = 12) +
	ylab("ECDF") +
	labs(group = "Distance function", col = "Distance function") +
	theme(legend.position = "bottom", legend.text = element_text(size = 14), legend.title.position = "top")
```

### 2. Test wether $G_j <= G_{ij}$

Next, we test if $G_j$ <= $G_{ij}$, which would imply a random sampling design. In that case, the algorithm would stop and return a random fold assignment. However, as can bee seen below (and also from the figure above), this is not the case:

```{r}
testks <- suppressWarnings(stats::ks.test(Gj, Gij, alternative = "great"))
testks$p.value >= 0.05
```


### 3. Cluster samples into $q$ clusters

Next, we cluster the training points into $q$ clusters, where $q \in Q$ and $Q = [k, ..., N]$. This means, that $q$ is a integer sequence ranging from the number of folds $k$ to the number of training points $N$. The length of this sequence is by default 100, but for visualization purposes it is reduced to 3 here.

```{r}
nk_length <- 3

clustgrid <- data.frame(
	nk = as.integer(round(exp(seq(
		log(k),
		log(nrow(samples) - 2),
		length.out = nk_length
	))))
)
clustgrid$W <- NA
clustgrid <- clustgrid[!duplicated(clustgrid$nk), ]
clustgroups <- list()

nk_df <- list()
tabclust <- list()
clust_nk <- list()
for (nk in clustgrid$nk) {
	# Create nk clusters
	clust_nk[[nk]] <- stats::kmeans(sample_coords, nk)$cluster
	# tabclust: table containing the unique clust_nks and their frequency
	tabclust[[nk]] <- as.data.frame(table(clust_nk[[nk]]))
	names(tabclust[[nk]]) <- c("clust_nk", "Freq")
	tabclust[[nk]]$clust_k <- NA
}

clustgrid
```

The sampled sequence of $q$ clusters is in this case `r clustgrid$nk`. The following figure shows the training points coloured by their cluster for each of the $q$ clusters:

```{r}
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| warning: false

for (nk in clustgrid$nk) {
	nk_df[[nk]] <- data.frame(
		"clust_nk" = clust_nk[[nk]],
		"nk" = rep(nk, length = length(clust_nk[[nk]])),
		"point_id" = samples$point_id
	)
}
nk_df_clean <- nk_df[lapply(nk_df, length) > 0]
nk_df_sf <- lapply(nk_df_clean, function(x) {
	coords <- st_coordinates(samples)
	df <- cbind(x, coords)
	df_sf <- st_as_sf(df, coords = c("X", "Y"), crs = st_crs(samples))
	df_sf
})

cluster_polys <- lapply(seq_along(nk_df_sf), function(i) {
	nk_df_focal <- nk_df_sf[[i]]

	nk_df_focal |>
		group_by(clust_nk, nk) |>
		summarise() |>
		st_convex_hull() |>
		mutate(clust_nk = as.factor(clust_nk), nk = nk)
})

lapply(seq_along(nk_df_sf), function(i) {
	pal <- colorspace::rainbow_hcl(length(unique(nk_df_sf[[i]]$clust_nk)))

	ggplot() +
		geom_sf(data = prediction_points, alpha = 0) +
		geom_sf(data = cluster_polys[[i]], aes(fill = as.factor(clust_nk)), alpha = 0.3, color = "transparent") +
		geom_sf(data = nk_df_sf[[i]], aes(colour = as.factor(clust_nk))) +
		scale_colour_manual(values = pal) +
		facet_wrap(~nk) +
		theme_bw(base_size = 8) +
		coord_sf(expand = FALSE, datum = st_crs(samples)) +
		theme(legend.position = "none", panel.grid = element_blank())
}) |>
	plot_grid(plotlist = _, nrow = 1)
```


### 4. Merge the $q$ clusters along their first PC into $k$ folds

The $q$ clusters are then merged along the first principal component of the samples coordinates if $q > k$. This prevents contigous clusters to end up in the same fold.

Therefore, we first calculate the PCA over the samples coordinates to capture the axis of largest spatial variation (first PC, red arrow):

```{r}
pcacoords <- stats::prcomp(sample_coords, center = TRUE, scale. = FALSE, rank = 1)
```

```{r}
#| fig-width: 8
#| fig-height: 4
#| echo: false

samples$pca <- pcacoords$x
center <- pcacoords$center
rotation_vector <- pcacoords$rotation[, 1]
arrow_length <- max(dist(sample_coords)) * 0.5
arrow_end <- center + arrow_length * rotation_vector
arrow_sf <- st_as_sf(
	data.frame(
		x = arrow_end[1],
		y = arrow_end[2]
	),
	coords = c("x", "y"),
	crs = st_crs(samples)
)

arrow_start <- st_as_sf(
	data.frame(
		x = center[1],
		y = center[2]
	),
	coords = c("x", "y"),
	crs = st_crs(samples)
)
ggplot() +
	geom_sf(data = prediction_points, alpha = 0) +
	geom_sf(data = samples, aes(colour = pca)) +
	geom_sf(data = arrow_start) +
	geom_segment(
		data = NULL,
		aes(x = center[1], y = center[2], xend = arrow_end[1], yend = arrow_end[2]),
		arrow = arrow(length = unit(0.35, "cm"), type = "closed"),
		linewidth = 1.2,
		color = "red"
	) +
	scale_colour_viridis_c("PC 1") +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	theme_bw(base_size = 16) +
	theme(panel.grid = element_blank())
```

Then, we order the $q$ clusters according to the first principal component:
```{r}
for (nk in clustgrid$nk) {
	# compute cluster centroids and project that centroid to the first PC
	centr_tpoints <- sapply(tabclust[[nk]]$clust_nk, function(x) {
		centrpca <- matrix(apply(sample_coords[clust_nk[[nk]] %in% x, , drop = FALSE], 2, mean), nrow = 1)
		colnames(centrpca) <- colnames(sample_coords)
		return(predict(pcacoords, centrpca))
	})
	# Order the clusters along this first PC to ensure spatial separation when merging them later
	tabclust[[nk]]$centrpca <- centr_tpoints
	tabclust[[nk]] <- tabclust[[nk]][order(tabclust[[nk]]$centrpca), ]
}
```

```{r}
#| fig-width: 8
#| fig-height: 4
#| echo: false

plot_nk <- list()
for (nk in clustgrid$nk) {
	plot_nk[[nk]] <- full_join(
		tabclust[[nk]] |> mutate("clust_nk" = as.factor(clust_nk)),
		nk_df[[nk]] |> mutate("clust_nk" = as.factor(clust_nk)),
		by = "clust_nk"
	)
}

plot_nk_clean <- plot_nk[lapply(plot_nk, length) > 0]
plot_nk_sf <- lapply(plot_nk_clean, function(x) {
	coords <- sample_coords |> as.data.frame()
	coords$point_id <- samples$point_id
	df <- full_join(x, coords, by = "point_id")
	df_sf <- st_as_sf(df, coords = c("X", "Y"), crs = st_crs(samples))
	df_sf
})

ggplot() +
	geom_sf(data = prediction_points, alpha = 0) +
	geom_sf(data = do.call("rbind", cluster_polys)) +
	geom_sf(data = do.call("rbind", plot_nk_sf), aes(colour = centrpca)) +
	scale_colour_viridis_c() +
	facet_wrap(~nk, nrow = 1) +
	theme_bw(base_size = 12) +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	theme(legend.position = "right", panel.grid = element_blank())
```

And lastly, we merge the $q$ clusters along the first principal component:

```{r}
clust_k <- list()
tabclust2 <- list()
for (nk in clustgrid$nk) {
	# We don't merge big clusters
	clust_i <- 1
	for (i in 1:nrow(tabclust[[nk]])) {
		if (tabclust[[nk]]$Freq[i] >= nrow(samples) / k) {
			tabclust[[nk]]$clust_k[i] <- clust_i
			clust_i <- clust_i + 1
		}
	}
	rm("clust_i")

	## And we merge the remaining nk clusters into k groups
	clust_i <- setdiff(1:k, unique(tabclust[[nk]]$clust_k))
	# tabclust is ordered along PC1 and thus repeating a sequence
	# from 1:k to assign folds results in spatially separated merging of clusters
	tabclust[[nk]]$clust_k[is.na(tabclust[[nk]]$clust_k)] <- rep(clust_i, ceiling(nk / length(clust_i)))[
		1:sum(is.na(tabclust[[nk]]$clust_k))
	]

	# Then creating a new data frame containing the raw nk cluster assignment for each point
	tabclust2[[nk]] <- data.frame(ID = 1:length(clust_nk[[nk]]), clust_nk = clust_nk[[nk]])
	# And add the assignment to k folds to them (clust_k)
	tabclust2[[nk]] <- merge(tabclust2[[nk]], tabclust[[nk]], by = "clust_nk")
	tabclust2[[nk]] <- tabclust2[[nk]][order(tabclust2[[nk]]$ID), ]
	clust_k[[nk]] <- tabclust2[[nk]]$clust_k
	tabclust2[[nk]]$nk <- nk # For plotting only
}
```

```{r}
#| fig-width: 8
#| fig-height: 4
#| echo: false

tabclust2_clean <- tabclust2[lapply(tabclust2, length) > 0]
tabclust2_sf <- lapply(tabclust2_clean, function(x) {
	coords <- st_coordinates(samples) |> as.data.frame()
	coords$point_id <- samples$point_id
	df <- full_join(x, coords, by = c("ID" = "point_id"))
	df_sf <- st_as_sf(df, coords = c("X", "Y"), crs = st_crs(samples))
	df_sf
})

cluster_polys2 <- lapply(seq_along(tabclust2_sf), function(i) {
	tabclust2_focal <- tabclust2_sf[[i]] |>
		st_drop_geometry() |>
		select(!ID) |>
		distinct() |>
		mutate(clust_nk = as.factor(clust_nk), clust_k = as.factor(clust_k))
	nk_df_sf[[i]] |>
		group_by(clust_nk) |>
		summarise() |>
		st_convex_hull() |>
		mutate(clust_nk = as.factor(clust_nk)) |>
		inner_join(y = tabclust2_focal, by = "clust_nk")
}) |>
	do.call(what = "rbind")

ggplot() +
	geom_sf(data = prediction_points, alpha = 0) +
	geom_sf(data = cluster_polys2, aes(fill = clust_k), alpha = 0.3, color = "transparent") +
	geom_sf(
		data = do.call("rbind", tabclust2_sf),
		aes(colour = as.factor(clust_k))
	) +
	ggthemes::scale_color_colorblind("k") +
	ggthemes::scale_fill_colorblind("k") +
	facet_wrap(~nk, nrow = 1) +
	theme_bw(base_size = 12) +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	theme(legend.position = "right", panel.grid = element_blank())
```

### 5. Calculate $G_j^*$ for each $q$ configuration and calculate $W$

```{r}
distclust_euclidean <- function(tr_coords, folds) {
	alldist <- rep(NA, length(folds))
	for (f in unique(folds)) {
		alldist[f == folds] <- c(FNN::knnx.dist(
			query = tr_coords[f == folds, , drop = FALSE],
			data = tr_coords[f != folds, , drop = FALSE],
			k = 1
		))
	}
	alldist
}

Gjstar_i <- list()
for (nk in clustgrid$nk) {
	# Compute W statistic if not exceeding maxp
	if (!any(table(clust_k[[nk]]) / length(clust_k[[nk]]) > 0.8)) {
		Gjstar_i[[nk]] <- distclust_euclidean(sample_coords, clust_k[[nk]])
		clustgrid$W[clustgrid$nk == nk] <- twosamples::wass_stat(Gjstar_i[[nk]], Gij)
		clustgroups[[paste0("nk", nk)]] <- clust_k[[nk]]
	}
}
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| echo: false

Gplot <- list()
for (nk in clustgrid$nk) {
	Gjstar_df <- data.frame(r = Gjstar_i[[nk]], Function = "2_Gjstar(r)")
	Gplot[[nk]] <- rbind(Gij_df, Gjstar_df, Gj_df)
	Gplot[[nk]]$nk <- nk
}

Gplot_df <- do.call("rbind", Gplot)

ggplot(
	data = Gplot_df,
	aes_string(x = "r", group = "Function", col = "Function")
) +
	geom_vline(xintercept = 0, lwd = 0.1) +
	geom_hline(yintercept = 0, lwd = 0.1) +
	geom_hline(yintercept = 1, lwd = 0.1) +
	stat_ecdf(geom = "step", lwd = 0.8) +
	scale_colour_manual(
		values = c(myColors[2], myColors[3], myColors[1]),
		labels = c("prediction-to-sample", "CV distances", "sample-to-sample")
	) +
	guides(colour = guide_legend(override.aes = list(linewidth = 1.6))) +
	theme_bw(base_size = 12) +
	ylab("ECDF") +
	labs(group = "Distance function", col = "Distance function") +
	theme(legend.position = "bottom", legend.text = element_text(size = 10), aspect.ratio = 1.6) +
	facet_wrap(~nk, nrow = 1)

clustgrid
```


### 6. Return the configuration yielding the lowest $W$
```{r}
k_final <- clustgrid$nk[which.min(clustgrid$W)]
W_final <- min(clustgrid$W, na.rm = T)
clust <- clustgroups[[paste0("nk", k_final)]]
Gjstar <- distclust_euclidean(sample_coords, clust)

print(k_final)
```

The following plot shows the training samples with their final fold assignment:

```{r}
#| fig-width: 8
#| fig-height: 4
#| echo: false

samples$fold <- clust

ggplot() +
	geom_sf(data = prediction_points, alpha = 0) +
	geom_sf(data = samples, aes(colour = as.factor(fold)), size = 2) +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	scale_color_brewer("Fold", palette = "Set1") +
	theme_bw() +
	theme(panel.grid = element_blank(), legend.position = "right") +
	labs(x = "", y = "")

```


## Summary

```{r}
#| fig-width: 12
#| fig-height: 14
#| warning: false
#| echo: false
lapply(seq_along(clustgrid$nk), function(i) {
	nk_focal <- clustgrid$nk[[i]]
	pal <- colorspace::rainbow_hcl(length(unique(nk_df_sf[[i]]$clust_nk)))

	tabclust2_focal <- tabclust2_sf[[i]] |>
		st_drop_geometry() |>
		select(!ID) |>
		distinct() |>
		mutate(clust_nk = as.factor(clust_nk), clust_k = as.factor(clust_k))

	cluster_polys <- nk_df_sf[[i]] |>
		group_by(clust_nk) |>
		summarise() |>
		st_convex_hull() |>
		mutate(clust_nk = as.factor(clust_nk)) |>
		inner_join(y = tabclust2_focal, by = "clust_nk")

	Nkplot_focal <- ggplot() +
		geom_sf(data = prediction_points, alpha = 0) +
		geom_sf(data = cluster_polys, aes(fill = clust_nk, colour = clust_nk)) +
		scale_fill_manual(values = pal) +
		scale_colour_manual(values = pal) +
		facet_wrap(~"Initial nk clusters") +
		theme_bw(base_size = 16) +
		coord_sf(expand = FALSE, datum = st_crs(samples)) +
		theme(legend.position = "none", panel.grid = element_blank())

	PCplot_focal <- ggplot() +
		geom_sf(data = prediction_points, alpha = 0) +
		geom_sf(data = cluster_polys, aes(fill = centrpca, colour = centrpca)) +
		scale_fill_viridis_c() +
		scale_colour_viridis_c() +
		facet_wrap(~"PCA", nrow = 1) +
		theme_bw(base_size = 16) +
		coord_sf(expand = FALSE, datum = st_crs(samples)) +
		theme(legend.position = "none", panel.grid = element_blank())

	Clustk_focal <- ggplot() +
		geom_sf(data = prediction_points, alpha = 0) +
		geom_sf(data = cluster_polys, aes(fill = clust_k, colour = clust_k)) +
		ggthemes::scale_fill_colorblind("clust_k") +
		ggthemes::scale_colour_colorblind("clust_k") +
		facet_wrap(~"Final k clusters", nrow = 1) +
		theme_bw(base_size = 16) +
		coord_sf(expand = FALSE, datum = st_crs(samples)) +
		theme(legend.position = "none", panel.grid = element_blank())

	Gplot_nkfocal <- ggplot(
		data = Gplot_df[Gplot_df$nk == nk_focal, ],
		aes_string(x = "r", group = "Function", col = "Function")
	) +
		geom_vline(xintercept = 0, lwd = 0.1) +
		geom_hline(yintercept = 0, lwd = 0.1) +
		geom_hline(yintercept = 1, lwd = 0.1) +
		stat_ecdf(geom = "step", lwd = 0.8) +
		scale_colour_manual(
			values = c(myColors[2], myColors[3], myColors[1]),
			labels = c("prediction-to-sample", "CV distances", "sample-to-sample")
		) +
		guides(colour = guide_legend(override.aes = list(linewidth = 1.6))) +
		theme_bw(base_size = 16) +
		ylab("") +
		labs(group = "Distance function", col = "Distance function") +
		theme(legend.position = "none", legend.text = element_text(size = 16)) +
		facet_grid(paste0("nk = ", nk) ~ "ECDF")

	plot_grid(Nkplot_focal, PCplot_focal, Clustk_focal, Gplot_nkfocal, nrow = 1, align = "vh")
}) |>
	plot_grid(plotlist = _, ncol = 1, align = "vh")
```