---
title: "k-NNDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k-NNDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Introduction
k-fold nearest-neighbour distance matching (kNNDM) cross-validation was developed by @Linnenbrink2024.
It is the k-fold version of NNDM (@Mila2022), and as such much faster and better suited towards medium to large sized datasets.
As NNDM, kNNDM also matches the distribution of nearest-neighbour distances (NND) between prediction locations and training samples during CV, but opposed to NNDM uses a clustering algorithm instead of excluding training samples.

The more detailed workflow is:

1. Conmpute the distribution of NNDs between training samples ($G_j$), as well as between prediction locations and training samples ($G_{ij}$).
2. Use the KS-test to evaluate wether $G_j >= G_{ij}$, i.e., wether the samples are randomly distributed in the prediction area.
3. A) If $G_j >= G_{ij}$; return a random CV split and stop the algorithm here.
3. B) Otherwise, continue by clustering the training points into $q \in Q$, where $Q = [k, ..., N]$. This yields a gradient from $k$ clusters (i.e., one cluster per fold) to $N$ clusters (i.e., one cluster per training sample).
4. The $q$ clusters are then merged along the first principal component of the training points coordinates' until $k$ is reached. This means that for $q = N$, nothing happens and a strongly clustered fold configuration is returned. For $q = N$, a random fold assignment is returned. For everything in-between ($q>k<N$), a gradient from strongly to weak clustered folds is returned.
5.  For each of the $q$ fold configurations, the NND distribution between the unique folds is calculated ($G_{j}^*$), and the distance between $G_j^*$ and $G_{ij}$ is calculated using the Wasserstein statistic (W).
6. The fold configuration yielding the lowest W is returned.


## Setup
```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
```

```{r setup}
library(PDAV)
library(terra)
library(sf)
library(simsam)
library(ggplot2)
library(tidyterra)
library(ggnewscale)
library(dplyr)
```

## Example data
### Simulate predictors and response
```{r}
r <- PDAV:::generate_rast()
samples <- PDAV:::generate_samples(r, 100)
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| echo: false

ggplot() +
	geom_spatraster(data = r) +
	coord_sf(expand = FALSE, datum = st_crs(r)) +
	scale_fill_viridis_c() +
	facet_wrap(~lyr) +
	theme_bw() +
	theme(panel.grid = element_blank())

ggplot() +
	geom_spatraster(data = r$outcome) +
	scale_fill_viridis_c("outcome") +
	new_scale_fill() +
	geom_sf(data = samples, shape = 21, aes(fill = "training samples"), size = 1) +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	scale_fill_manual("", values = "white") +
	facet_wrap(~sampling) +
	theme_bw() +
	theme(panel.grid = element_blank())
```

## kNNDM CV

```{r}
#| warning: false

predictor_stack <- r[[setdiff(names(r), "outcome")]]

results <- lapply(unique(samples$sampling), function(smpling) {
	samples_smpl <- samples |>
		filter(sampling == smpling) |>
		select(!sampling)

	cate_rasters <- which(names(r) %in% c("forest", "grass"))

	knndm
})

names(results) <- unique(samples$sampling)
```