---
title: "NNDM"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{NNDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
Nearest-neighbour distance matching (NNDM) leave-one-out (LOO) cross-validation was developed by @Mila2022.
It matches the distribution of nearest-neighbour distances (NNDs) encountered during CV to the one encountered during prediction.
This is achieved by excluding a variable number of the nearest training points of each left-out point from the training set. 
Because NNDM is LOO-based, it requires as many model fits as there are training points.
It is therefore more flexible, but also considerably slower, than kNNDM. NNDM is best suited to small datasets, and kNNDM should be preferred for larger ones.

Three NND distributions are involved:

- *Ĝ*<sub>ij</sub>: the empirical distribution of distances from each prediction point to its nearest training point;
- *Ĝ*<sub>j</sub>: the empirical distribution of distances from each training point to its nearest other training point, i.e. the NND distribution of plain LOO CV;
- *Ĝ*<sub>j</sub><sup>\*</sup>: the modified version of *Ĝ*<sub>j</sub> produced by NNDM, which is matched to *Ĝ*<sub>ij</sub>.

The more detailed workflow is:
<ol> 
<li>Compute <em>Ĝ<sub>ij</sub></em> and <em>Ĝ<sub>j</sub></em>.</li> 
<li>Choose <em>φ</em>, the distance up to which matching is performed. By default (<code>phi = "max"</code>) the maximum observed distance is used, so matching covers the entire range. Alternatively, an estimate of the autocorrelation range can be supplied; beyond <em>φ</em> no training points are excluded and all data are used for training.</li> 
<li>Derive <em>Ĝ<sub>j</sub><sup>*</sup></em>, starting from plain LOO CV (<em>Ĝ<sub>j</sub><sup>*</sup></em> = <em>Ĝ<sub>j</sub></em>) and processing all training-point NNDs in increasing order. For the current distance <em>r</em>, which belongs to left-out point <em>j</em> and its nearest remaining training point <em>k</em>:
 <ol type="a"> 
 <li>check whether removing <em>k</em> from the training set of <em>j</em> would still leave <em>Ĝ<sub>j</sub><sup>*</sup></em>(<em>r</em>) ≥ <em>Ĝ<sub>ij</sub></em>(<em>r</em>), i.e. whether CV distances are still shorter than prediction distances at <em>r</em>;
 </li> <li>check whether fold <em>j</em> would still retain at least <code>min_train</code> (default 50%) of the training data;
 </li> <li>if both conditions hold, exclude <em>k</em> from fold <em>j</em> and update <em>Ĝ<sub>j</sub><sup>*</sup></em>; otherwise leave fold <em>j</em> unchanged. Exclusions are never undone;</li>
 <li>move on to the next-smallest NND, and stop once it exceeds <em>φ</em>.</li> 
 </ol> 
 </li> <li>Return, for each left-out point, the indices of the test point, the retained training points and the excluded points (<code>indx_test</code>, <code>indx_train</code>, <code>indx_exclude</code>). These can be passed directly to caret's <code>trainControl(index = , indexOut = )</code> or used to set up a custom resampling strategy in mlr3.</li> 
 </ol>


## Setup
```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
```

```{r setup}
library(PDAV)
library(terra)
library(sf)
library(simsam)
library(ggplot2)
library(cowplot)
library(tidyterra)
library(ggnewscale)
library(dplyr)

set.seed(100)

k <- 2
```

## Example data
### Simulate predictors and response
```{r}
r <- PDAV:::generate_rast()
samples <- PDAV:::generate_samples(r, 100) |>
	filter(sampling == "biased")
samples$point_id <- 1:nrow(samples)
samples <- select(samples, point_id)
```

```{r}
#| fig-width: 8
#| fig-height: 4
#| echo: false
ggplot() +
	geom_spatraster(data = r) +
	coord_sf(expand = FALSE, datum = st_crs(r)) +
	scale_fill_viridis_c() +
	facet_wrap(~lyr) +
	theme_bw() +
	theme(panel.grid = element_blank())

ggplot() +
	geom_spatraster(data = r$outcome) +
	scale_fill_viridis_c("outcome") +
	new_scale_fill() +
	geom_sf(data = samples, shape = 21, aes(fill = "training samples"), size = 2) +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	scale_fill_manual("", values = "white") +
	theme_bw() +
	theme(panel.grid = element_blank())
```



## NNDM in geographical space


### 1. Compute G<sub>j</sub> and G<sub>ij</sub>

Firstly, the NNDM algorithm calculates the distribution of NND between samples (G<sub>j</sub>), as well as the distribution of NNDs between prediction points and samples (G<sub>ij</sub>). The calculation of NNDs is shown in the following figure: the nearest neighbour distance between samples is shown in the left panel, while the right panel shows nearest neighbour distances between prediction points and samples.

```{r}
#| fig-width: 8
#| fig-height: 4
#| echo: false
prediction_points <- spatSample(r, size = 1000, method = "regular", as.points = TRUE) |>
	st_as_sf()

pred_coords <- st_coordinates(prediction_points)
sample_coords <- st_coordinates(samples)

# Gj
idx_nn_s <- st_nearest_feature(samples)
nearest_samples_s <- samples[idx_nn_s, ]
samp_coords_s <- st_coordinates(nearest_samples_s)
lines_s <- mapply(
	function(i) {
		st_linestring(rbind(sample_coords[i, ], samp_coords_s[i, ]))
	},
	seq_len(nrow(sample_coords)),
	SIMPLIFY = FALSE
)
lines_sf_s <- st_sf(
	geometry = st_sfc(lines_s, crs = st_crs(prediction_points))
)

# Gij
idx_nn <- st_nearest_feature(prediction_points, samples)
nearest_samples <- samples[idx_nn, ]
samp_coords <- st_coordinates(nearest_samples)
lines <- mapply(
	function(i) {
		st_linestring(rbind(pred_coords[i, ], samp_coords[i, ]))
	},
	seq_len(nrow(pred_coords)),
	SIMPLIFY = FALSE
)
lines_sf <- st_sf(
	geometry = st_sfc(lines, crs = st_crs(prediction_points))
)


NND_samples <- ggplot() +
	geom_sf(data = prediction_points, alpha = 0) +
	geom_sf(data = samples, shape = 4) +
	geom_sf(data = lines_sf_s, linewidth = 0.2, color = "black") +
	scale_shape_manual("", values = c(1, 4)) +
	coord_sf(expand = FALSE, datum = st_crs(prediction_points)) +
	ggtitle("NND between training points") +
	theme_bw() +
	theme(legend.position = "None", panel.grid = element_blank(), plot.title = element_text(size = 12))

NND_pred_samples <- ggplot() +
	geom_sf(data = prediction_points, aes(shape = "prediction points"), color = "lightblue") +
	geom_sf(data = samples, aes(shape = "training points")) +
	geom_sf(data = lines_sf, linewidth = 0.2, color = "lightblue") +
	scale_shape_manual("", values = c(1, 4)) +
	coord_sf(expand = FALSE, datum = st_crs(prediction_points)) +
	ggtitle("NND between prediction and training points") +
	theme_bw() +
	theme(legend.position = "None", panel.grid = element_blank(), plot.title = element_text(size = 12))

plot_grid(NND_samples, NND_pred_samples, nrow = 1, align = "vh")
```

To characterize the distribution of these NNDs, their empirical cumulative density functions (ECDFs) are calculated:

```{r}
# Compute nearest neighbour distances between training and prediction points
Gij <- sf::st_distance(prediction_points, samples)
units(Gij) <- NULL
Gij <- apply(Gij, 1, min)

# Compute distance matrix of training points
tdist <- sf::st_distance(samples)
units(tdist) <- NULL
diag(tdist) <- NA
Gj <- apply(tdist, 1, function(x) min(x, na.rm = TRUE))
```

```{r}
#| fig-width: 4
#| fig-height: 4
#| echo: false
#| warning: false
myColors <- RColorBrewer::brewer.pal(3, "Dark2")

Gij_df <- data.frame(r = Gij[order(Gij)])
Gij_df$Function <- "1_Gij(r)"
Gj_df <- data.frame(r = Gj[order(Gj)])
Gj_df$Function <- "3_Gj(r)"

# Merge data for plotting
Gplot <- rbind(Gij_df, Gj_df)

ggplot(
	data = Gplot,
	aes(x = r, group = Function, colour = Function)
) +
	geom_vline(xintercept = 0, lwd = 0.1) +
	geom_hline(yintercept = 0, lwd = 0.1) +
	geom_hline(yintercept = 1, lwd = 0.1) +
	stat_ecdf(geom = "step", lwd = 0.8) +
	scale_colour_manual(
		values = c(myColors[2], myColors[1]),
		labels = c("prediction-to-training", "training-to-training")
	) +
	guides(colour = guide_legend(override.aes = list("linewidth" = 1.6))) +
	theme_bw(base_size = 12) +
	ylab("ECDF") +
	labs(group = "Distance function", col = "Distance function") +
	theme(legend.position = "bottom", legend.text = element_text(size = 14), legend.title.position = "top")
```


### 2. Calculate G<sub>j</sub><sup>*</sup> for each left-out point starting from LOO-CV

Secondly, the algorithm starts with the smallest NND found among the training points, i.e. with the left-out point j whose nearest training point k is closest. It then tests whether excluding k from the training set of j would keep Ĝ<sub>j</sub><sup>*</sup>(r) ≥ Ĝ<sub>ij</sub>(r) at that distance r -- that is, whether CV distances are still shorter than prediction distances there -- and whether fold j would still retain at least min_train (50% by default) of the training data. If both hold, k is excluded from that fold only; it remains available for training in all other folds. The algorithm then moves on to the next-smallest NND, which usually belongs to a different left-out point. Otherwise the exclusion is skipped and the algorithm proceeds to the next-larger NND. This continues until the current distance exceeds φ.

```{r}
# initialize Gjstar
Gjstar <- Gj

min_train <- 0.5
dimnames(tdist) <- NULL
tdist0 <- tdist # keep an untouched copy for plotting

# select the held-out point (jmin) with the closest distance to its nearest training point (kmin)
rmin <- min(Gjstar)
jmin <- which.min(Gjstar)[1]
kmin <- which(tdist[jmin, ] == rmin)

# set phi to the maximum relevant distance
phi <- max(c(Gij, c(tdist)), na.rm = TRUE) + 1e-9

# Check if removing the point yields a higher Gjstar than the target Gij
# Target Gij (at the current distance):
current_Gij <- (sum(Gij <= rmin) / length(Gij))
# Gjstar that would result from excluding the training point:
current_Gjstar <- (sum(Gjstar <= rmin) - 1) / length(Gjstar) # -1: if kmin were excluded, the NND of point jmin would exceed rmin,so it would no longer be counted
Gjstar_greater_check <- current_Gjstar >= current_Gij

# Check if the minimum number of training data is preserved:
min_train_check <- sum(!is.na(tdist[jmin, ])) / ncol(tdist) > min_train

# for plotting only: create a list that captures the output of the algorithm
snap <- list(list(j = NA_integer_, k = integer(0), r = NA_real_, Gjstar = Gjstar))

# If both checks pass, proceed with the NNDM algorithm:
while (rmin <= phi) {
	if (
		(sum(Gjstar <= rmin) - 1) / length(Gjstar) >= sum(Gij <= rmin) / length(Gij) &&
			sum(!is.na(tdist[jmin, ])) / ncol(tdist) > min_train
	) {
		tdist[jmin, kmin] <- NA
		Gjstar <- apply(tdist, 1, min, na.rm = TRUE)
		snap[[length(snap) + 1L]] <- list(j = jmin, k = kmin, r = rmin, Gjstar = Gjstar)

		rmin <- min(Gjstar[Gjstar >= rmin])
	} else if (sum(Gjstar > rmin) == 0) {
		break
	} else {
		rmin <- min(Gjstar[Gjstar > rmin])
	}
	jmin <- which(Gjstar == rmin)[1]
	kmin <- which(tdist[jmin, ] == rmin)
}

indx_exclude <- lapply(seq_len(nrow(tdist)), \(i) setdiff(which(is.na(tdist[i, ])), i))
```

The following Figure shows the sequential removal of points in the NNDM algorithm. It starts from plain LOO CV (left) to the completed matching (right).
In each column the algorithm is processing the current nearest-neighbour distance `r` (dashed line in the bottom row): the left-out point is blue, the neighbour excluded from its fold orange, and faint grey lines show all pairs excluded earlier. The blue point differs between columns because distances are processed in increasing order across all folds, not fold by fold. `r` grows from left to right, so the orange link becomes longer. In the bottom row Ĝ<sub>j</sub><sup>*</sup> is adjusted from short NNDs resulting from plain LOO CV towards the target Ĝ<sub>ij</sub>. The right panel is the last iteration that produced an exclusion that was accepted (the maximum distance allowed by φ would have been `r round(phi)`, but the maximum exclusion distance was `r round(max(snap[[length(snap)]]$Gjstar))`).

```{r}
#| fig-width: 8
#| fig-height: 5.6
#| out-width: "100%"
#| echo: false

sel <- c(1L, round(length(snap) / 2), length(snap)) # LOO CV + 2 stages

lvls <- vapply(
	seq_along(sel),
	function(i) {
		t <- sel[i]
		if (t == 1L) {
			"LOO CV (no exclusions)"
		} else if (t == length(snap)) {
			sprintf("final state (%d exclusions)", t - 1L)
		} else {
			sprintf("after %d exclusions", t - 1L)
		}
	},
	character(1)
)

xy <- st_coordinates(samples)
seg <- function(t) st_linestring(rbind(xy[as.integer(snap[[t]]$j), ], xy[as.integer(snap[[t]]$k)[1], ]))

## ---- row 1: maps ----------------------------------------------------
pts_df <- do.call(
	rbind,
	lapply(seq_along(sel), function(i) {
		t <- sel[i]
		d <- samples
		d$status <- "training point"
		if (t > 1L) {
			d$status[as.integer(snap[[t]]$k)] <- "excluded point"
			d$status[as.integer(snap[[t]]$j)] <- "left-out test point"
		}
		d$col <- factor(lvls[i], levels = lvls)
		d
	})
)

## pairs excluded in earlier iterations (faint grey)
hist_df <- do.call(
	rbind,
	lapply(seq_along(sel), function(i) {
		ts <- setdiff(seq_len(sel[i] - 1L), 1L)
		if (!length(ts)) {
			return(NULL)
		}
		st_sf(col = factor(lvls[i], levels = lvls), geometry = st_sfc(lapply(ts, seg), crs = st_crs(samples)))
	})
)

## pair excluded in the current iteration (bold orange)
now_df <- do.call(
	rbind,
	lapply(seq_along(sel), function(i) {
		if (sel[i] == 1L) {
			return(NULL)
		}
		st_sf(col = factor(lvls[i], levels = lvls), geometry = st_sfc(list(seg(sel[i])), crs = st_crs(samples)))
	})
)

p_map <- ggplot() +
	{
		if (!is.null(hist_df)) geom_sf(data = hist_df, colour = "grey55", linewidth = 0.2, alpha = 0.25)
	} +
	{
		if (!is.null(now_df)) geom_sf(data = now_df, colour = "#E66101", linewidth = 0.9)
	} +
	geom_sf(data = pts_df, aes(colour = status, size = status)) +
	facet_wrap(~col, nrow = 1) +
	scale_colour_manual(
		"",
		values = c("training point" = "grey78", "excluded point" = "#E66101", "left-out test point" = "#2166AC")
	) +
	scale_size_manual(
		values = c("training point" = 1.1, "excluded point" = 3, "left-out test point" = 3.4),
		guide = "none"
	) +
	guides(colour = guide_legend(override.aes = list(size = 3))) +
	coord_sf(expand = FALSE, datum = NA) +
	theme_bw(base_size = 12) +
	theme(
		panel.grid = element_blank(),
		axis.text = element_blank(),
		axis.ticks = element_blank(),
		strip.text = element_text(size = 11),
		plot.margin = margin(2, 2, 2, 2)
	)

## ---- row 2: ECDFs ---------------------------------------------------
ecdf_df <- do.call(
	rbind,
	lapply(seq_along(sel), function(i) {
		rbind(
			data.frame(r = Gij, curve = "Gij(r)", col = lvls[i]),
			data.frame(r = snap[[sel[i]]]$Gjstar, curve = "Gj*(r)", col = lvls[i])
		)
	})
)
ecdf_df$col <- factor(ecdf_df$col, levels = lvls)

vl <- data.frame(col = factor(lvls, levels = lvls), r = vapply(sel, function(t) snap[[t]]$r, numeric(1)))

p_ecdf <- ggplot(ecdf_df, aes(x = r, colour = curve)) +
	geom_vline(data = vl, aes(xintercept = r), linetype = "dashed", colour = "grey40", na.rm = TRUE) +
	stat_ecdf(geom = "step", lwd = 0.8) +
	facet_wrap(~col, nrow = 1) +
	coord_cartesian(xlim = c(0, stats::quantile(Gij, 0.97))) +
	scale_colour_manual(
		"",
		values = c("Gij(r)" = myColors[1], "Gj*(r)" = myColors[2]),
		labels = c("prediction-to-training", "training-to-training")
	) +
	guides(colour = guide_legend(override.aes = list(linewidth = 1.5))) +
	labs(x = "r", y = "ECDF") +
	theme_bw(base_size = 12) +
	theme(strip.text = element_blank(), panel.grid = element_blank(), plot.margin = margin(2, 2, 2, 2))

## ---- assemble with one shared legend row ----------------------------
leg <- function(p) {
	cowplot::get_plot_component(
		p + theme(legend.position = "bottom", legend.margin = margin(0, 0, 0, 0)),
		"guide-box-bottom",
		return_all = TRUE
	)
}

legends <- plot_grid(leg(p_map), leg(p_ecdf), nrow = 1)

plot_grid(
	plot_grid(
		p_map + theme(legend.position = "none"),
		p_ecdf + theme(legend.position = "none"),
		ncol = 1,
		align = "v",
		axis = "lr",
		rel_heights = c(1, 1)
	),
	legends,
	ncol = 1,
	rel_heights = c(1, 0.10)
)
```