---
title: "Prediction-domain adaptive evaluation"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Prediction-domain adaptive evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette provides an overview over the different methods of prediction-domain adaptive evaluation. First, I will briefly summarize the (dis-)advantages of the different methods (based on my subjective impression), and then I will briefly show how they work using a simulated case study with three different sampling designs. For a more detailed description of the methods, please refer to the dedicated vignettes.

## Overview of the currently developed methods:

| Method | Advantages | Disadvantages | Critical Parameters |
| ----------- | ----------- | ----------- | ----------- |
| [NNDM](NNDM.html)| - **direct quantification of the prediction situation** through nearest-neighbour distances<br>- **NND plots** reveal situations where prediction situation cannot be matched during CV<br>- works natively with **distances in the geographic or feature space**  | - **high computational costs** since it uses LOO-CV<br>- **May remove large fractions of the training samples** and thus lead to unstable models<br>- depends on the estimated autocorrelation range (though not so important as in spatial+ CV used in DA-CV) | - `phi`: Autocorrelation threshold up to which Nearest-Neighbour distances are matched<br>- `min_train`: fold balancing |
| [kNNDM](k-NNDM.html) | - **direct quantification of the prediction situation** through nearest-neighbour distances<br>- **NND plots** reveal situations where prediction situation cannot be matched during CV<br>- more computational efficient than NNDM<br>- works natively with **distances in the geographic or feature space** | - **less flexible** in matching the NND distributions than NNDM | - `maxp`: maximum fold size allowed: higher numbers lead to more imbalanced splits, but also may resemble the prediction situation better |
| [DA-CV](DA-CV.html)| - creates a **map that depicts areas of inter- vs extrapolation**, which can be useful for subsequent sampling or for uncertainty assessment | - **less direct** approximation of the prediction situation<br>- No inspection method available to **assess if the prediction situation can be matched** during CV<br>- Matches the prediction situation in the feature space, but **mixes spaces** in the spatial+ resampling method<br>- requires calculating the weighted **average of two validation statistics**, which might be problematic for metrics other than RMSE | - `autoc_threshold`: block size in spatial cross-validation |
| [TWCV](twcv.html) | - applies **weighting** to match the prediction situation, which is more flexible than resampling-based matching of the prediction situation | - **integration of geographic distances** implicit and weighting of the predictor vs geographical space is automatically done by the raking algorithm<br>- is **prone to extrapolation** (raking fails when it encounteres strong extrapolation)<br>- Depends on the resampling method (e.g., random resampling might not create geographical distances that are comparable to the prediction situation) | - resampling method |


## Case Study
```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
```

```{r setup}
library(PDAV)
library(dplyr)
library(caret)
library(terra)
library(sf)
library(simsam)
library(ggplot2)
library(cowplot)
library(tidyterra)
library(ggnewscale)

set.seed(100)
```

### Simulate predictors and response
```{r}
r <- PDAV:::generate_rast()
predictor_stack <- r[[setdiff(names(r), "outcome")]]
cate_rasters <- which(names(r) %in% c("forest", "grass"))

samples <- PDAV:::generate_samples(r, 100)
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| echo: false
ggplot() +
	geom_spatraster(data = r) +
	coord_sf(expand = FALSE, datum = st_crs(r)) +
	scale_fill_viridis_c() +
	facet_wrap(~lyr) +
	theme_bw() +
	theme(panel.grid = element_blank())
```


```{r}
#| fig-width: 6
#| fig-height: 4
#| echo: false
ggplot() +
	geom_spatraster(data = r$outcome) +
	scale_fill_viridis_c("outcome") +
	new_scale_fill() +
	geom_sf(data = samples, shape = 21, aes(fill = "training samples"), size = 1) +
	coord_sf(expand = FALSE, datum = st_crs(samples)) +
	scale_fill_manual("", values = "white") +
	facet_wrap(~sampling) +
	theme_bw() +
	theme(panel.grid = element_blank())
```

### DA-CV

DA-CV (@Wang2025) uses adversial validation (AV) to predict the probability that a prediction location is similar to the training samples.
Then it calculates a RMSE based on random CV, as well as spatial+ CV (@Wang2023) and weights both of them according to the relative area of similar cells (random CV) and dissimilar cells (spatial+ CV). 

```{r}
#| warning: false
results <- lapply(unique(samples$sampling), function(smpling) {
	samples_smpl <- samples |>
		filter(sampling == smpling) |>
		select(!sampling)

	da_cv(
		samples = samples_smpl,
		predictors = r,
		response = "outcome",
		folds_k = 5,
		autoc_threshold = 1
		#cate_col_start = min(cate_rasters),
		#cate_col_end = max(cate_rasters)
	)
})

names(results) <- unique(samples$sampling)
```

```{r da-cv-values, include=FALSE}
auc_random <- round(results$random$auc, 1)

auc_clustered <- round(results$clustered$auc, 2)
D_clustered <- (auc_clustered - 0.5) / 0.5
T_clustered <- 0.5 * D_clustered

w_similar <- round(results$clustered$weights[["similar"]], 2)
w_dissimilar <- round(results$clustered$weights[["different"]], 2)
```

For randomly distributed samples, the AV classifier has a performance of AUC = `r auc_random`.
This is then normalized to D = 0, because the classifier is not better than random guessing at distinguishing whether a prediction location is similar or dissimilar to the training samples; see @Wang2025b, section 2.1, for more information.
The threshold is then calculated as T(D) = 0.5 × 0, which is 0.
This means that all prediction locations are considered similar to the training samples, and no extrapolation correction is required, as shown in the following map.
Consequently, all weight is assigned to random CV.

For clustered samples, the AV classifier achieves a performance of AUC = `r auc_clustered`.
This leads to D = (`r auc_clustered` - 0.5) / (1 - 0.5) = `r round(D_clustered, 2)`.
The threshold is then T(D) = 0.5 × `r round(D_clustered, 2)` = `r round(T_clustered, 2)`.
Hence, all prediction cells with a similarity score lower than `r round(T_clustered, 2)` are classified as dissimilar.
The relative fraction of prediction locations classified as similar to the sampling locations is `r w_similar`, while the fraction classified as dissimilar is `r w_dissimilar`.

``` {r}
#| fig-width: 12
#| fig-height: 5
#| fig-cap: "Figure 3: Similarity maps resulting from the AV classifier."
lapply(unique(samples$sampling), function(smpling) {
	plot(results[[smpling]]) +
		new_scale_fill() +
		geom_sf(data = samples[samples$sampling == smpling, ], shape = 21, fill = "white") +
		coord_sf(expand = FALSE, datum = st_crs(samples[samples$sampling == smpling, ], )) +
		ggtitle(smpling) +
		theme(
			legend.position = "right",
			plot.title = element_text(face = "bold", hjust = 0.5)
		)
}) |>
	plot_grid(plotlist = _, nrow = 1, align = "vh")
```

Lastly, the resulting cross-validation fold assignments can be used to calculate the weighted RMSE of a spatial predictive model.

```{r}
form <- as.formula(paste0("outcome~", paste0(names(predictor_stack), collapse = "+")))
pgrid <- data.frame(mtry = 6, splitrule = "variance", min.node.size = 5)

weighted_RMSE <- lapply(unique(samples$sampling), function(smpling) {
	samples_smpl <- samples |>
		filter(sampling == smpling) |>
		mutate("randomCV" = results[[smpling]]$folds_RDM, "spatialCV" = results[[smpling]]$folds_SP) |>
		st_drop_geometry()

	folds_random <- CAST::CreateSpacetimeFolds(samples_smpl, spacevar = "randomCV", k = 5)
	folds_spatial <- CAST::CreateSpacetimeFolds(samples_smpl, spacevar = "spatialCV", k = 5)

	train_cntrl_random <- trainControl(
		method = "CV",
		index = folds_random$index,
		indexOut = folds_random$indexOut,
		savePredictions = TRUE
	)

	train_cntrl_SP <- trainControl(
		method = "CV",
		index = folds_spatial$index,
		indexOut = folds_spatial$indexOut,
		savePredictions = TRUE
	)

	rand_mod <- train(
		form,
		data = samples_smpl,
		method = "ranger",
		trControl = train_cntrl_random,
		tuneGrid = pgrid
	)

	spat_mod <- train(
		form,
		data = samples_smpl,
		method = "ranger",
		trControl = train_cntrl_SP,
		tuneGrid = pgrid
	)

	err_stats_rand <- CAST::global_validation(rand_mod)
	err_stats_SP <- CAST::global_validation(spat_mod)

	err_stats_weighted <- sqrt(
		results[[smpling]]$weights[["similar"]] *
			(err_stats_rand[["RMSE"]]^2) +
			results[[smpling]]$weights[["different"]] * (err_stats_SP[["RMSE"]]^2)
	)

	prediction <- predict(r, rand_mod)
	list(err_stats_weighted, prediction)
})
```

```{r}
#| fig-width: 6
#| fig-height: 8
#| echo: false
samples_long <- lapply(unique(samples$sampling), function(smpling) {
	samples_smpl <- samples |>
		filter(sampling == smpling) |>
		mutate("randomCV" = results[[smpling]]$folds_RDM, "spatialCV" = results[[smpling]]$folds_SP) |>
		select(randomCV, spatialCV, sampling)
	samples_smpl$ID <- 1:nrow(samples_smpl)
	samples_geom <- select(samples_smpl, ID)
	st_drop_geometry(samples_smpl) |>
		pivot_longer(-c(ID, sampling)) |>
		left_join(samples_geom, by = "ID") |>
		st_as_sf() |>
		mutate(
			weight = case_when(
				name == "randomCV" ~ round(results[[smpling]]$weights[["similar"]], 2),
				name == "spatialCV" ~ round(results[[smpling]]$weights[["different"]], 2)
			)
		)
}) |>
	do.call(what = "rbind")

ggplot() +
	geom_sf(data = samples_long, aes(colour = as.factor(value)), size = 1) +
	geom_text(
		data = samples_long,
		aes(x = 45, y = 180, label = paste0("Weight: ", weight)),
		size = 3,
		fontface = "bold",
		check_overlap = TRUE
	) +
	coord_sf(expand = FALSE, datum = st_crs(samples_long)) +
	facet_grid(sampling ~ name) +
	scale_color_brewer("Fold", palette = "Set1") +
	theme_bw() +
	theme(panel.grid = element_blank(), legend.position = "right") +
	labs(x = "", y = "")
```

The RMSE obtained by DA-CV are `r round(weighted_RMSE[[1]][[1]], 3)` for the random sampling design, `r round(weighted_RMSE[[2]][[1]], 3)` for the biased sampling and `r round(weighted_RMSE[[3]][[1]], 3)` for the clustered design.


### kNNDM CV

```{r}

```


### NNDM CV

### TWCV