---
title: "TWCV"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{TWCV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
Target-Weighted Cross-Validation (TWCV) was developed by @Brenning2026. 
It uses iterative calibration weighting to align the validation task (predictor values of the training points and nearest-neighbour distances (NNDs) between folds) to the prediction task (predictor values of the prediction points and NNDs between prediction and training points).

The more detailed workflow is:

1. Compute CV losses by training a model on the training folds fold and predicting to the held-out fold
2. Compute the CV validation task and the prediction task (NNDs and predictor values)
3. Calculate quantiles of the balancing variables for training and prediction points separately (NNDs and predictors)
4. Calculate the proportion of training points and prediction points falling in each quantile
5. Applies iterative proportional fitting ("raking") to align the proportions of the training points to those in the prediction points:
   1. For each balancing variable, calculate the current number of training points in each quantile.
   2. Calculate the target number of training points in each quantile based on the target proportions.
   3. Calculate multiplicative adjustment factors that transform the current counts to the target counts.
   4. Because the weighting is done for each variable seperately ("marginal"), correcting one variable can disturb variables adjusted earlier. 
   5. Repeat this process until the maximum relative change in weights between successive iterations falls below a threshold.
6. Normalize weights and shrink them towards 1 to mitigate extreme values.


## Setup
```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
```

```{r setup}
library(PDAV)
library(dplyr)
library(caret)
library(terra)
library(sf)
library(simsam)
library(ggplot2)
library(cowplot)
library(tidyterra)
library(ggnewscale)

set.seed(100)
```

## Simulate predictors and response
```{r}
r <- PDAV:::generate_rast()
predictor_stack <- r[[setdiff(names(r), "outcome")]]
cate_rasters <- which(names(r) %in% c("forest", "grass"))

n_sample <- 100

samples <- sam_field(
	x = r,
	size = n_sample,
	method = sample_clustered(nclusters = 10, radius = 30, na.rm = TRUE)
)

predictor_vars <- names(predictor_stack)
env_vars <- predictor_vars
task_vars <- env_vars
response <- "outcome"
pred_points <- spatSample(predictor_stack, 1000, method = "regular", na.rm = TRUE, as.points = TRUE) |>
	st_as_sf()

sample_coords <- st_coordinates(samples) |> as.data.frame() |> rename("x" = X, "y" = Y)
grid_coords <- st_coordinates(pred_points) |> as.data.frame() |> rename("x" = X, "y" = Y)
sample_dat <- st_drop_geometry(samples) |>
	mutate(id = row_number()) |>
	cbind(sample_coords)
grid_dat <- st_drop_geometry(pred_points) |>
	mutate(id = row_number()) |>
	cbind(grid_coords)

twcv_specs <- PDAV:::make_twcv_specs(
	predictor_vars = predictor_vars,
	include_distance = TRUE,
	balance_by = 0.2,
	shrink_lambda = 0.2,
	name = "twcv_extended"
)

fit_fun <- function(train_dat, model, response, predictor_vars = NULL, ...) {
	if (is.null(predictor_vars)) {
		predictor_vars <- setdiff(names(train_dat), response)
	}

	dat <- train_dat[, c(response, predictor_vars), drop = FALSE]

	ranger::ranger(
		formula = stats::reformulate(predictor_vars, response = response),
		data = dat,
		...
	)
}

predict_fun <- function(fit, newdata, ...) {
	predict(fit, data = newdata)$predictions
}

model = "rf"

folds <- sample(rep(1:5, length.out = n_sample))
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Figure 1: The predictor stack"
#| echo: false
ggplot() +
	geom_spatraster(data = predictor_stack) +
	coord_sf(expand = FALSE, datum = st_crs(predictor_stack)) +
	scale_fill_viridis_c() +
	facet_wrap(~lyr) +
	theme_bw() +
	theme(panel.grid = element_blank())
```


## Run TWCV

### 1. Compute CV losses

Iterate over each fold and calculate the predictions and the resulting pointwise CV error:
```{r}
pred <- PDAV:::compute_cv_predictions(
	sample_dat = sample_dat,
	folds = folds,
	model = model,
	response = response,
	fit_fun = fit_fun,
	predict_fun = predict_fun
)

pe <- PDAV:::compute_pointwise_errors(
	obs = sample_dat[[response]],
	pred = pred
)

cv_losses <- data.frame(
	id = sample_dat$id,
	fold = folds,
	obs = pe$obs,
	pred = pe$pred,
	error = pe$error,
	se = pe$se,
	ae = pe$ae
)

knitr::kable(head(cv_losses))
```

### 2. Compute the CV validation task and the prediction task

```{r}
d_sample <- PDAV:::nearest_neighbor_distance(
	query_coords = sample_dat[, c("x", "y")],
	ref_coords = sample_dat[, c("x", "y")],
	exclude_self = TRUE
)

d_grid <- PDAV:::nearest_neighbor_distance(
	query_coords = grid_dat[, c("x", "y")],
	ref_coords = sample_dat[, c("x", "y")],
	exclude_self = FALSE
)

sample_tasks <- data.frame(
	id = sample_dat$id,
	d = d_sample,
	stringsAsFactors = FALSE
)

grid_tasks <- data.frame(
	id = grid_dat$id,
	d = d_grid,
	stringsAsFactors = FALSE
)

for (v in env_vars) {
	sample_tasks[[v]] <- sample_dat[[v]]
	grid_tasks[[v]] <- grid_dat[[v]]
}

tasks <- list(
	sample_tasks = sample_tasks,
	grid_tasks = grid_tasks
)

sample_desc <- tasks$sample_tasks
idx <- match(cv_losses$id, sample_desc$id)

# combine CV losses with task descriptors for training points
out <- cv_losses
for (v in task_vars) {
	if (!(v %in% names(sample_desc))) {
		stop("Variable '", v, "' missing in sample_tasks.", call. = FALSE)
	}
	out[[v]] <- sample_desc[[v]][idx]
}

# Calculates the NND between folds based on distance matrix / matrices
d_realized <- PDAV:::compute_cv_prediction_distances(
	sample_dat = sample_dat,
	folds = cv_losses$fold
)

# Update the NND attached to the training points from NND between points to NND between folds
idx_d <- match(out$id, sample_dat$id)
out$d <- d_realized[idx_d]

aug <- list(
	losses = out,
	grid_tasks = tasks$grid_tasks
)

knitr::kable(head(aug$losses))
knitr::kable(head(aug$grid_tasks))
```


### 3. Calculate quantiles of the balancing variables

```{r}
cv_losses <- aug$losses
grid_tasks <- aug$grid_tasks

unweighted_losses <- list(
	unweighted = PDAV:::summarize_losses(cv_losses)
)

sample_tasks_bal <- PDAV:::prepare_for_balancing(
	df = cv_losses,
	vars = twcv_specs$twcv_extended$balancing_vars,
	ref_df = grid_tasks,
	by = twcv_specs$twcv_extended$balance_by
)

grid_tasks_bal <- PDAV:::prepare_for_balancing(
	df = grid_tasks,
	vars = twcv_specs$twcv_extended$balancing_vars,
	ref_df = grid_tasks,
	by = twcv_specs$twcv_extended$balance_by
)

bal <- list(
	sample_tasks_bal = sample_tasks_bal,
	grid_tasks_bal = grid_tasks_bal
)

knitr::kable(head(bal$sample_tasks_bal))
knitr::kable(head(bal$grid_tasks_bal))
```

```{r plot, echo=FALSE, fig.width=10, fig.height=5}
#-----------------------------
# 1. Helper: order Q-levels
#-----------------------------

get_q_levels <- function(x) {
	if (is.factor(x)) {
		levs <- levels(x)
	} else {
		levs <- unique(as.character(stats::na.omit(x)))
	}

	qnum <- suppressWarnings(as.integer(sub(".*_Q", "", levs)))

	if (all(!is.na(qnum))) {
		levs[order(qnum)]
	} else {
		sort(levs)
	}
}

#-----------------------------
# 2. Create bands from existing quintile variable
#-----------------------------

make_bands_from_existing_bins <- function(ref_dat, x_var = "elev", cat_var = "elev_cat", xlim = NULL) {
	levs <- get_q_levels(ref_dat[[cat_var]])

	ranges <- ref_dat %>%
		transmute(
			x = .data[[x_var]],
			cat = factor(as.character(.data[[cat_var]]), levels = levs)
		) %>%
		filter(!is.na(x), !is.na(cat)) %>%
		group_by(cat) %>%
		summarise(
			lo = min(x),
			hi = max(x),
			.groups = "drop"
		) %>%
		arrange(cat)

	# Boundaries between adjacent existing bins.
	# Since exact cut points are not stored, these are reconstructed
	# from adjacent observed bin ranges.
	cuts <- (head(ranges$hi, -1) + tail(ranges$lo, -1)) / 2

	if (is.null(xlim)) {
		xlim <- range(ref_dat[[x_var]], na.rm = TRUE)
	}

	tibble(
		cat = ranges$cat,
		xmin = c(xlim[1], cuts),
		xmax = c(cuts, xlim[2])
	)
}

#-----------------------------
# 3. Common x-range across both data frames
#-----------------------------

x_rng <- range(
	c(
		bal$sample_tasks_bal$elev,
		bal$grid_tasks_bal$elev
	),
	na.rm = TRUE
)

#-----------------------------
# 4. Build elevation bands using existing grid elev_cat
#-----------------------------

elev_bands <- make_bands_from_existing_bins(
	ref_dat = bal$grid_tasks_bal,
	x_var = "elev",
	cat_var = "elev_cat",
	xlim = x_rng
)

#-----------------------------
# 5. Plot function
#-----------------------------

make_density_band_plot <- function(dat, bands, title = "", x_var = "elev", x_lab = "Elevation") {
	ggplot(dat, aes(x = .data[[x_var]])) +
		geom_rect(
			data = bands,
			aes(
				xmin = xmin,
				xmax = xmax,
				ymin = -Inf,
				ymax = Inf,
				fill = cat
			),
			inherit.aes = FALSE,
			alpha = 0.18
		) +
		geom_density(
			fill = "grey70",
			color = "grey20",
			alpha = 0.35,
			linewidth = 1
		) +
		geom_vline(
			data = bands[-nrow(bands), ],
			aes(xintercept = xmax),
			inherit.aes = FALSE,
			linetype = "dashed",
			color = "grey30"
		) +
		coord_cartesian(
			xlim = range(c(bands$xmin, bands$xmax), na.rm = TRUE)
		) +
		scale_fill_brewer(
			palette = "Set2",
			drop = FALSE
		) +
		labs(
			title = title,
			x = x_lab,
			y = "Density",
			fill = "Elevation quintile"
		) +
		theme_minimal() +
		theme(
			legend.position = "bottom"
		)
}

#-----------------------------
# 6. Create individual plots
#-----------------------------

p_sample <- make_density_band_plot(
	dat = bal$sample_tasks_bal,
	bands = elev_bands,
	title = "Sample tasks",
	x_var = "elev",
	x_lab = "Elevation"
)

p_grid <- make_density_band_plot(
	dat = bal$grid_tasks_bal,
	bands = elev_bands,
	title = "Grid tasks",
	x_var = "elev",
	x_lab = "Elevation"
)

#-----------------------------
# 7. Extract shared legend
#-----------------------------

shared_legend <- cowplot::get_legend(
	p_sample +
		theme(
			legend.position = "bottom",
			legend.box = "horizontal"
		)
)

# Remove legends from individual panels
p_sample_noleg <- p_sample + theme(legend.position = "none")
p_grid_noleg <- p_grid + theme(legend.position = "none")

#-----------------------------
# 8. Combine panels and shared legend
#-----------------------------

combined_plot <- cowplot::plot_grid(
	cowplot::plot_grid(
		p_sample_noleg,
		p_grid_noleg,
		nrow = 1,
		align = "hv",
		axis = "tblr",
		labels = c("A", "B")
	),
	shared_legend,
	ncol = 1,
	rel_heights = c(1, 0.12)
)

combined_plot
```

### 4. Calculate frequencies for the quantiles

```{r}
balance_df <- as.data.frame(
	lapply(twcv_specs$twcv_extended$balancing_vars, function(v) sample_tasks_bal[[paste0(v, "_cat")]])
)
names(balance_df) <- twcv_specs$twcv_extended$balancing_vars

# Calculates proportion of predpoints in each quantile of each predictor used for weighting
target_margins <- PDAV:::compute_target_margins_generic(
	grid_tasks_bal = grid_tasks_bal,
	balancing_vars = twcv_specs$twcv_extended$balancing_vars
)
target_margins$elev
```

### 5. Apply Raking

```{r, message=FALSE, warning=FALSE, results='hide'}
margin_names <- names(target_margins)
max_iter <- 500
tol = 1e-6
n <- nrow(balance_df)
w <- rep(1, n)

for (iter in seq_len(max_iter)) {
	w_old <- w

	for (m in margin_names) {
		x <- as.integer(balance_df[[m]])
		levs <- seq_along(target_margins[[m]])
		target_prop <- target_margins[[m]]

		# calculate the number of training points currently in each quantile
		# uses the weights from the previous balancing variable
		# -> already weighted, but likely not ideally for this variable
		current_totals <- tapply(w, factor(x, levels = levs), sum)
		current_totals[is.na(current_totals)] <- 0

		# calculate the desired number of training points in each quantile that matches the target margins:
		# Number of training points * target proportion vector
		target_totals <- sum(w) * target_prop

		adj <- rep(1, length(levs))
		ok <- current_totals > 0

		# calculates the weight needed to adjust the training point distribution to the target margins
		# (weights > 1 are used to up-weigh underrepresented classes, weights < 1 to down-weight over-represented ones)
		adj[ok] <- target_totals[ok] / current_totals[ok]
		adj[!ok] <- NA_real_

		valid <- !is.na(adj[x])
		w[valid] <- w[valid] * adj[x[valid]]
		print(paste(m, w))
	}

	# Compute the relative strength of the absolute change of weights from base (or previous) to new weights
	rel_change <- max(abs(w - w_old) / pmax(abs(w_old), 1e-12))

	# When the changes converge (i.e., when the weight difference from previous iteration to new one is small), stop and return weights
	# (when the weights for predictor B are changed, weights for predictor A might be off again, and another iteration starts,
	# until the diff between them is small)
	if (rel_change < tol) break
}

converged <- iter < max_iter || rel_change < tol

tw <- list(
	weights = w,
	converged = converged,
	iterations = iter
)
```

```{r}
tw$weights
```

### 6. Normalize and shrink weights

```{r}
# Weights are normalized by their mean and shrinked towards 1 to mitigate extreme values
shrink_lambda <- 0
tw$weights_raw <- PDAV:::normalize_weights(tw$weights)
tw$weights <- PDAV:::shrink_weights(tw$weights_raw, lambda = shrink_lambda)
tw$shrink_lambda <- twcv_specs$twcv_extended$shrink_lambda
tw$balancing_vars <- twcv_specs$twcv_extended$balancing_vars

tw$weights
```

### 7. Return weighted error estimates

```{r}
est_list <- PDAV:::summarize_losses(cv_losses, tw$weights)
weight_objects <- tw

result <- list(
	losses = cv_losses,
	estimators = est_list,
	weights = weight_objects,
	twcv_specs = twcv_specs
)

# Unweighted error:
unweighted_losses

# Weighted error:
result$estimators
```