---
title: "DA-CV"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{DA-CV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
Target-Weighted Cross-Validation (TWCV) was developed by @Brenning2026. 
It uses iterative calibration weighting to align the validation task (predictor values of the training points and nearest-neighbour distances (NNDs) between folds) to the prediction task (predictor values of the prediction points and NNDs between prediction and training points).

The more detailed workflow is:

1. Compute CV losses by training a model on the training folds fold and predicting to the held-out fold
2. Compute the CV validation task and the prediction task (NNDs and predictor values)
3. Calculates quantiles of the balancing variables for training and prediction points separately (NNDs and predictors)
4. Calculates the proportion of training points and prediction points falling in each quantile
5. Applies iterative proportional fitting ("raking") to align the proportions of the training points to those in the prediction points:
   1. For each balancing variable, calculate the current number of training points in each quantile.
   2. Calculate the target number of training points in each quantile based on the target proportions.
   3. Calculate multiplicative adjustment factors that transform the current counts to the target counts.
   4. Because the weighting is done for each variable seperately ("marginal"), correcting one variable can disturb variables adjusted earlier. 
   5. Repeat this process until the maximum relative change in weights between successive iterations falls below a threshold.
6. Normalize weights and shrink them towards 1 to mitigate extreme values.


## Setup
```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>"
)
```

```{r setup}
library(PDAV)
library(dplyr)
library(caret)
library(terra)
library(sf)
library(simsam)
library(ggplot2)
library(cowplot)
library(tidyterra)
library(ggnewscale)

set.seed(100)
```

## Simulate predictors and response
```{r}
r <- PDAV:::generate_rast()
predictor_stack <- r[[setdiff(names(r), "outcome")]]
cate_rasters <- which(names(r) %in% c("forest", "grass"))

samples <- sam_field(
	x = r,
	size = 100,
	type = "clustered",
	type_opts = list(nclusters = 10, radius = 30),
	na.rm = TRUE
) |>
	terra::extract(x = r, bind = TRUE) |>
	st_as_sf()
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Figure 1: The predictor stack"
#| echo: false
ggplot() +
	geom_spatraster(data = predictor_stack) +
	coord_sf(expand = FALSE, datum = st_crs(predictor_stack)) +
	scale_fill_viridis_c() +
	facet_wrap(~lyr) +
	theme_bw() +
	theme(panel.grid = element_blank())
```


## Run DA-CV
